Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
Thursday PM editing pass.

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
        per generation, respectively.
        For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
        \label{fig:stats_performance}
    }
\end{figure}

@@

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is this population genetics statistic summarising?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
essentially uses the signal from nearby patterns of variation
to interpret relationships at a given site,
thus tranforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
In this paper, we have made an exploratory foray in this direction,
by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
between several subsets of the 1000 Genomes dataset,
in several inferred tree sequences.
It is not yet clear whether these time profiles tell us more about history
or about biases of the the inference methods,
but either result would be useful.

@@

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing backmutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
would give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

A final crucial assumption underlying this duality is that mutations can be treated
as indepenent of the trees themselves.
This is clearly not always true --- for instance,
a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
but seems likely to be approximately true.
\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
but more work needs to be done, particularly on models with large proprotions of sites under selection.

@@

The statistics defined here are all \emph{additive} along the genome:
if we have computed a statistic in equal-size windows along a chromsome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
We have also defined site statistics to be additive across \emph{alleles},
to allow the definitions to apply to sites with more than two alleles.
This decision has some nice properties, but is not the only reasonable choice.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgements}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
and to Wilder Wohns, Yan Wong and Leo Speidel for providing inferred tree sequences.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} GEVA, and
        \textbf{(bottom right)} Relate.
        \label{fig:chr20_full}
    }
\end{figure}

>>>>

\krt{I have related concerns. What does it mean to look at a branch stat and a site state
    from an inferred tree sequence?  The input genotype data are informative about both 
    the inferred edges and their branch lengths.   So it seems to me that comparing the two types of stats is comparing
    non-independent quantities, and I am unclear how to interpret that? (I admit to not studying the GEVA paper closely
yet, nor the Speidel).  Presumably, the site stat is going to give you the right value (e.g., match whatever the
original paper reported for, say, pi), but the branch version of the stat could differ from this for perhaps a lot 
of reasons that all contribute to "edge estimation error"?  Not calling variants below a certain frequency,
preferentially calling variants you've seen in a previous study, or imputing in variants from other studies?
Overall, I have to squint at Fig S2 to see differences, and most of the issue seems to be near what I assume to be a
centromere?}

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
        per generation, respectively.
        For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
        algorithm used by \texttt{tskit}. 
        \label{fig:stats_performance}
    }
\end{figure}

@@

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is this population genetics statistic summarising?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
essentially uses the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
In this paper, we have made an exploratory foray in this direction,
by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
between several subsets of the 1000 Genomes dataset,
in several inferred tree sequences.
It is not yet clear whether these time profiles tell us more about history
or about biases of the inference methods,
but either result would be useful.

@@

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing back mutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
would give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

A final crucial assumption underlying this duality is that mutations can be treated
as independent of the trees themselves.
This is clearly not always true --- for instance,
a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
but seems likely to be approximately true.
\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
but more work needs to be done, particularly on models with large proportions of sites under selection.

@@

The statistics defined here are all \emph{additive} along the genome:
if we have computed a statistic in equal-size windows along a chromosome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
We have also defined site statistics to be additive across \emph{alleles},
to allow the definitions to apply to sites with more than two alleles.
This decision has some nice properties, but is not the only reasonable choice.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgments}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
and to Wilder Wohns, Yan Wong and Leo Speidel for providing inferred tree sequences.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} GEVA, and
        \krt{This should be tsinfer+GEVA, no?}
        \textbf{(bottom right)} Relate.
        \label{fig:chr20_full}
    }
\end{figure}
